---
output:
  html_document: default
  pdf_document: default
  always_allow_html: true
---

```{r setup, include=FALSE} 
knitr::opts_chunk$set(warning = FALSE, message = FALSE) 
```

# **Cyclistic Analysis**

## Set up
Load of libraries:
```{r, error=FALSE, warning=FALSE, message=FALSE}

library("tidyverse")
library("lubridate")
library(dplyr)
library(knitr)
library(geosphere)
library(mapview)
```

Here we load all the different CSV into a single data frame.
```{r, error=FALSE, warning=FALSE, message=FALSE}
df <- list.files(path = "./dataset", pattern = "*.csv", full.names = TRUE) %>% 
  lapply(read_csv) %>%                                          
  bind_rows 
```


First we want to see some general stats of the data.
```{r}
head(df)
```

```{r}
summary(df)
```

## Data aggregation
To better understand the data and arrive at a conclusion, we add some new columns.
```{r}
df$ride_length <- df$ended_at - df$started_at
df$day_of_week <- wday(df$started_at)
df$year <- year(df$started_at)
df$month <- month(df$started_at)
df$day <- day(df$started_at)
df$hour <- hour(df$started_at)

df$ride_distance <- distGeo(matrix(c(df$start_lng, df$start_lat), ncol = 2), matrix(c(df$end_lng, df$end_lat), ncol = 2))
df$ride_distance <- df$ride_distance / 1000

```

## Consistency of features
For features like rideable_type and member_casual, we can check their unique values 
to verify if there are some strange labels. But as we can see, the values seem correct.

```{r}
unique(df$rideable_type)
unique(df$member_casual)
```
As ride_id must be unique, we can check for duplicated values, but none was found.
```{r}
n_occur <- data.frame(table(df$ride_id))
n_occur[n_occur$Freq > 1,]
```

Check the length for ride_id, it is always 16.
```{r}
unique(nchar(as.character(df$ride_id)))
```

Wrong data as rides shorter than a minute have been dropped from the dataset.
```{r}
df <- subset(df, ride_length > 60)
```

## Type transformation
Features like member_casual can be represented as a boolean type, so we can map 
this column and rename it.
```{r}
df$is_member <- df$member_casual == 'member'
df$member_casual <- NULL
```


## Dealing with null values
Null values:
```{r}
colSums(is.na(df))
```
Dropping nulls found:
```{r}
df <- subset(df, !is.na(df$start_station_id))
df <- subset(df, !is.na(df$end_station_id))
df <- subset(df, !is.na(df$start_station_name))
df <- subset(df, !is.na(df$end_station_name))
df <- subset(df, !is.na(df$end_lat))
df <- subset(df, !is.na(df$end_lng))
```


## Analysis through plots

Comparative of members and casual riders. As we can see, there are slightly more subscribed clients than casuals. 
```{r}
ggplot(df, aes(is_member, fill=is_member)) +
    geom_bar() +
    labs(x="Casuals and Members", y="Count", title="Subscription distribution")
```

```{r}
df %>% 
    group_by(is_member) %>% 
    summarize(count = n())
```

We can check which type of client does more kilometers:
```{r}
df %>% 
    group_by(is_member) %>% 
    summarize(mean = mean(ride_distance))
```


### Distribution of rides through time.

```{r}
df %>%
  ggplot(aes(month, fill=is_member)) +
    geom_bar() +
    labs(x="Month", y="Rides", title="Rides per month")
```
\

The months with more activity are the ones from May to October, this could be due to the good weather. It is also important to notice that casual riders are much more sensitive to this characteristic.

```{r}
df %>%
  ggplot(aes(day_of_week, fill=is_member)) +
    geom_bar() +
    labs(x="Day of week", y="Rides", title="Rides per day of week")
```
\

As per the day of the week, Saturdays and Sundays are slightly more active than work days. We can see that on work days, there are many more members than casual riders. This can give us the hypothesis that clients that need to ride bikes between Monday to Friday (work, school...) found more attractive the membership status. And casual riders are more concentrated through the weekend.\

In addition, we can observe that on these days, the average time of the ride is bigger than the other days. People spend more time cycling at the weekend:

```{r}
df %>% 
    group_by(day_of_week) %>% 
    summarize(mean = mean(ride_length))
```



```{r}
df %>%
    ggplot(aes(day, fill=is_member)) +
    labs(x="Day", y="Rides", title="Rides per day") +
    geom_bar()
```
\

If we inspect the days of the month with more rides, we can see that more or less the distribution is stable through all the days except the 31, but that is because half of the months don't have this day.

```{r}
df %>%
    ggplot(aes(hour, fill=is_member)) +
    labs(x="Hour", y="Count", title="Rides per hour") +
    geom_bar()
```


The hours with more activity are from 8 AM to 7 PM, and there are important spikes with a high presence of members around those two hours, coinciding with the rush hours from work or school.
\

We can increase our scope in a hypothetical ad campaign if we focus the resources on those hours, and we can check as well the top 10 used ride stations, where we could arrive to more potential clients:
```{r}
stations_rides <- df %>% 
  group_by(start_station_name) %>% 
  summarize(rides_per_station = n()) %>%
  arrange(desc(rides_per_station))

head(stations_rides, 10)
```

As we have two types of bicycles (plus the docked ones), we can plot the distribution:

```{r}
ggplot(df, aes(rideable_type, fill=is_member)) +
    geom_bar() +
    labs(x="Ride type", y="Count", title="Distribution of ride types")
```
\

Classic bikes are much more used than electric ones. And in this table, we can see the time spent on each type. Docked bikes are the most used, but this is because when they are docked the time still runs until they are parked at a proper station. Rides with classic bikes tend to be longer:

```{r}
df %>% 
    group_by(rideable_type) %>% 
    summarize(mean = mean(ride_length))
```

Rides with electric bikes are not only 21% faster, as we can see in the following table, but they also do an average of an extra 400m compared with clients that stick with classic bikes:

```{r}
df %>% 
    group_by(rideable_type) %>% 
    summarize(mean = mean(ride_distance))
```

As for the length of rides, members spend an average of 13 min for each ride and casual riders 28 min:


```{r}
df_avg_length_rides <- df %>% 
    group_by(is_member) %>% 
    summarize(avg_ride_length_mins = mean(as.numeric(ride_length, units="mins"))) 

df_avg_length_rides
```

Which are the favorite destinations of our clients? On the following map, we indicate the top 5 destination stations:
```{r}

stations_rides <- df %>% 
  group_by(end_station_name) %>% 
  summarize(count = n()) %>%
  arrange(desc(count))
head(stations_rides, 10)

first_10 <- head(stations_rides, 10)


```

For a more visual experience, we can plot each station on a map, mapping its importance with the size of dots:
```{r}
require(data.table)
setDT(df); setDT(first_10) # convert to data.tables by reference

# join to get GPS coord from the top end stations with the number of rides
join <- df[first_10, mult = "first", on = "end_station_name", nomatch=0L] 
join$rides_per_station <- first_10$count
# selection of the features to plot
df_end_stations <- select(join, c("end_station_name", "end_lng", "end_lat","rides_per_station"))
# resize for point size on map
df_end_stations$rides_per_station <- (df_end_stations$rides_per_station / max(df_end_stations$rides_per_station))*25
# map plot
mapview(df_end_stations, xcol = "end_lng", ycol = "end_lat", crs = 4269, 
        cex = df_end_stations$rides_per_station, grid = FALSE, layer.name = 'Top destination stations')
```
It seems that the Navy Pier is the most visited area of Chicago.

But we can also check the different behavior between members and casual riders. 
Here we see the most used pair of stations (start and end) for members: 
```{r}
df_members <- filter(df, is_member == TRUE)

df_members$comb_station_names <- paste(df_members$start_station_name, " - " ,df_members$end_station_name)
tail(names(sort(table(df_members$comb_station_names))), 5)
```

And the same with casual riders:
```{r}
df_casuals <- filter(df, is_member == FALSE)

df_casuals$comb_station_names <- paste(df_casuals$start_station_name, " - " ,df_casuals$end_station_name)
tail(names(sort(table(df_casuals$comb_station_names))), 5)
```



