# **Cyclistic Analysis**

## Set up
Load of libraries:
```{r, error=FALSE, warning=FALSE, message=FALSE}

library("tidyverse")
library("lubridate")
library(dplyr)
library(knitr)
library(geosphere)
library(mapview)
```

Here we load all the different csv into a single data frame.
```{r, error=FALSE, warning=FALSE, message=FALSE}
df <- list.files(path = "./dataset", pattern = "*.csv", full.names = TRUE) %>% 
  lapply(read_csv) %>%                                          
  bind_rows 
```
 
```{r}
unique_gps_coords <- unique(df[c("start_lat", "start_lng")])
```





```{r}
sub_sample <- df[sample(nrow(df), 100000), ]
```






First we want to see some general stats of the data.
```{r}
head(df)
```

```{r}
summary(df)
```

## Data aggregation
To better understand the data and arrive to a conclusion, we add some new columns.
```{r}
df$ride_length <- df$ended_at - df$started_at
df$day_of_week <- wday(df$started_at)
df$year <- year(df$started_at)
df$month <- month(df$started_at)
df$day <- day(df$started_at)
df$hour <- hour(df$started_at)

df$ride_distance <- distGeo(matrix(c(df$start_lng, df$start_lat), ncol = 2), matrix(c(df$end_lng, df$end_lat), ncol = 2))
df$ride_distance <- df$ride_distance / 1000

```

## Consistency of features
For features like rideable_type and member_casual, we can check its unique values 
to verify if there are some strange labels. But as we can see, the values seems correct.

```{r}
unique(df$rideable_type)
unique(df$member_casual)
```
As ride_id must be unique, we can check for duplicated value, but none was found.
```{r}
n_occur <- data.frame(table(df$ride_id))
n_occur[n_occur$Freq > 1,]
```

Check the length for ride_id, it is always 16.
```{r}
unique(nchar(as.character(df$ride_id)))
```

Wrong data as rides shorter than a minute have been dropped from the dataset.
```{r}
df <- subset(df, ride_length > 60)
```

## Type transformation
Features like member_casual can be represented as a boolean type, so we can map 
this column an rename it.
```{r}
df$is_member <- df$member_casual == 'member'
df$member_casual <- NULL
```


## Dealing with null values
Null values:
```{r}
colSums(is.na(df))
```
Dropping nulls found:
```{r}
df <- subset(df, !is.na(df$start_station_id))
df <- subset(df, !is.na(df$end_station_id))
df <- subset(df, !is.na(df$start_station_name))
df <- subset(df, !is.na(df$end_station_name))
df <- subset(df, !is.na(df$end_lat))
df <- subset(df, !is.na(df$end_lng))
```


## Analysis through plots

Comparative of members and casual riders. As we can see, there are slightly more subscribed clients than casuals. 
```{r}
ggplot(df, aes(is_member, fill=is_member)) +
    geom_bar() +
    labs(x="Casuals and Members", y="Count", title="Subscription distribution")
```

```{r}
df %>% 
    group_by(is_member) %>% 
    summarize(count = n())
```

We can check which type of client do more kilometers:
```{r}
df %>% 
    group_by(is_member) %>% 
    summarize(mean = mean(ride_distance))
```
Both ride a similar distance, non members ride an average of 200m more. And those who choose to ride an electric bike do an extra 300m in average than the clients that stick with the classic bike:


```{r}
df %>% 
    group_by(rideable_type) %>% 
    summarize(mean = mean(ride_distance))
```


### Distribution of rides through time.

```{r}
df %>%
  ggplot(aes(month, fill=is_member)) +
    geom_bar() +
    labs(x="Month", y="Rides", title="Rides per month")
```
\

The months with more activity are the ones from May to October, this could be due to the good weather. It is also important to notice that casual riders are much more sensible to this characteristic.

```{r}
df %>%
  ggplot(aes(day_of_week, fill=is_member)) +
    geom_bar() +
    labs(x="Day of week", y="Rides", title="Rides per day of week")
```
\

As per the day of week, Saturdays and Sundays are slightly more active than work days. We can see that on work days, there are much more members than casual riders. This can give us the hypothesis that clients that need to ride bikes between Monday to Friday (work, school...) found more attractive the membership status. And casual riders are more concentrate through the weekend.  

\ 

In addition, We can observe that in this days, the average time of the ride is bigger that the other days. People spend more time cycling at weekend:

```{r}
df %>% 
    group_by(day_of_week) %>% 
    summarize(mean = mean(ride_length))
```



```{r}
df %>%
    ggplot(aes(day, fill=is_member)) +
    labs(x="Day", y="Rides", title="Rides per day") +
    geom_bar()
```
\

If we inspect the days of the month with more rides, we can see that more or less the distribution is stable through all the days except the 31, but that is because half of the months don't have this day.

```{r}
df %>%
    ggplot(aes(hour, fill=is_member)) +
    labs(x="Hour", y="Count", title="Rides per hour") +
    geom_bar()
```


The hours with more activity are from 8AM to 7PM, and there are important spikes with a high presence of members around those two hours, coinciding with the rush hours from work or school.
\

We can increase our scope in an hypothetical ad campaign if we focus the resources on those hours, and we can check as well the top 10 used ride stations, where we could arrive to more potential clients:
```{r}
stations_rides <- df %>% 
  group_by(start_station_name) %>% 
  summarize(rides_per_station = n()) %>%
  arrange(desc(rides_per_station))

head(stations_rides, 10)
```

As we have two types of bicycles (plus the docked ones), we can plot the distribution:

```{r}
ggplot(df, aes(rideable_type, fill=is_member)) +
    geom_bar() +
    labs(x="Ride type", y="Count", title="Distribution of ride types")
```
\

Classic bikes are much more used than the electric ones. And in this table we can see that TODO
```{r}
df %>% 
    group_by(rideable_type) %>% 
    summarize(mean = mean(ride_length))
```


\ 

As for the length rides, casual riders tend to spend more time than members. Here we can see the difference for each hour:

```{r, }
ggplot(data=df) +
  geom_bar(mapping = aes(x=hour, y=ride_length, fill=is_member), stat='identity') + 
  labs(x="Hour", y="Ride length", title="Ride length per hour")
```


In general, members spend an average of 13 min for each ride and casual riders 28 min:


```{r}
df_avg_length_rides <- df %>% 
    group_by(is_member) %>% 
    summarize(avg_ride_length_mins = mean(as.numeric(ride_length, units="mins"))) 

df_avg_length_rides
```


```{r}
first_10 <- head(df, 10)
last_10 <- tail(df, 10)

mapview(first_10, xcol = "start_lng", ycol = "start_lat", crs = 4269, grid = FALSE)

```
Most frequent start and end station for member riders:
```{r}
df_members <- filter(df, is_member == TRUE)

df_members$comb_station_names <- paste(df_members$start_station_name, " - " ,df_members$end_station_name)
tail(names(sort(table(df_members$comb_station_names))), 5)
```

Most frequent start and end station for casual riders:
```{r}
df_casuals <- filter(df, is_member == FALSE)

df_casuals$comb_station_names <- paste(df_casuals$start_station_name, " - " ,df_casuals$end_station_name)
tail(names(sort(table(df_casuals$comb_station_names))), 5)
```



